Numerical Investigation of Viscous Flow Instabilities in Multiphase Heterogeneous Porous Media

### 2018-01-16 Papers/ViscousInstabilities/Manuscript_ViscousInstabilities_KC.tex
updates on abstract etc.

\begin{abstract} 
  The aim of this paper is to model flow instabilities \ie viscous fingering using a novel method for modeling multi-phase flows in heterogeneous porous media. Starting from modelling the Hele-Shaw experimental apparatus, which is mathematically analogous to the two-dimensional flow in porous media problem. The development of finger structures is investigated by taking into account different conditions such as, different domains under uniform or different permeabilities regions, mobility ratios, mesh types. A novel control finite element method (CVFEM) is used to discretize the governing equations that will solve the multi-fluid porous media flow model. The governing multiphase porous media flow equations are solved using an unstructured mesh by introducing adaptivity in later stages. Dynamic mesh adaptivity will not only help to better capture the development of finger but also allow the system to focus the computational effort where is needed, minimizing both the CPU time and the numerical diffusion. It was found that mobility ratio along with the initial and boundary conditions, as well as the use of adaptivity on top of an ustructured mesh allow the user to better describe and visualise the phenomenon.

\red{KOSTAS: Abstracts are divided into 4 parts: (1) Motivation (1-2 sentences); (2) Main aim of the manuscript highlighting the novelty of the work (you must not use the word novelty ... this should be implicit, 2-6 sentences); (3) summary of the main technologies/methods (2-5 sentences) and; (4) summary of the main findings (1-3 sentences).}     
\end{abstract}

@@

\medskip
%Due to the heterogeneity of geological formations, detailed modelling and simulation of multiphase flows through porous media is crucial for design, control and prediction of the aforementioned processes. 
Description of physics and mechanisms of multiphase porous media flows were reported by \citet{wooding_1976} with focus on capillary pressure and flow regimes. A comprehensive review of force balances at the interface between immiscible fluids and resulting mechanisms for flow instabilities can be found in \citet{homsy_1987}. Flow instabilities can be classified as macroscopic and microscopic, the former is due to imposed boundary conditions, whereas the later is associated to local phenomena at the fluids interface \citep[\eg Kelvin-Helmholtz and Saffman-Taylor instabilities][]{saffman_1959}. This work focuses on flow instabilities in two-phase systems due to viscous and stress forces (often referred as viscous instabilities or viscous fingering, Section \ref{section:ViscousInstabilities}).
%Description of physics and mechanisms of multiphase flows were reported by \citet{wooding_1976} with focus on capillary pressure and flow regimes. A comprehensive review of mechanisms of fluid instabilities (\ie fingering) and forces acting on the interface between two immiscible fluids can be found in \citet{homsy_1987}. \citet{adler_1988} described both microscopic and macroscopic scale Darcy-equations.
% Nevertheless we should refer to the area of flow instabilities  as these, first described by \citet{saffman_1958} \red{(IS THIS TRUE? SAFFMANN MAY HAVE BEEN THE FIRST TO THROUGHLY DISCUSS VISCOUS INSTABILITIES, BUT THERE WERE OTHER SCIENTISTS THAT DESCRIBED FLUID INSTABILITIES BEFORE, E.G., RAYLEIGH(1896), TAYLOR (1955) ETC)}. 
%The first distinction should be made between microscopic and macroscopic instabilities. Microscopic instabilities describe phenomena occurring locally, for example, Kelvin-Helmholtz and Saffman-Taylor instabilities \citep{saffman_1959}. On the other hand, macroscopic instabilities involve two-phase flow system \citep{bottoni_1992}. This paper focuses on microscopic instabilities in two-phase systems that occurs due to viscous and stress forces, often referred as viscous instabilities or \textit{viscous fingering} (see section \ref{section:ViscousInstabilities}).

@@

\medskip
The displacement efficiency depends upon the ratio of the viscous to capillary forces or capillary number. 
%Flow instabilities will occur due to: pressure gradient along the interface of fluids; stress forces acting in opposite directions; mobility differences that may exceed critical limits and; heterogeneities of the domain (layering). 
When viscous force of the injected fluid overcomes the capillary force, hydrodynamic instabilities may occur, resulting in the collapse of the interface between fluids and start forming fingers.

\medskip
Viscous flow instabilities can be found across several disciplines and scales, from chemical separation processes to geological reservoir fluids. \citet{muskat_1934} investigated fluid flow in Hele-Shaw cells, \ie parallel flat plates separated by an infinitesimal gap, and the impact on the capillary number ($N_{c}$),
\begin{equation}
N_{c} = \frac{\mu U}{\gamma}.
\label{eq:capillary_number}
\end{equation}
In Eqn.~\ref{eq:capillary_number}, $\gamma$ is the surface tension and $U$ is the characteristic velocity of the moving interface. This experimental apparatus enables instabilities to be qualitatively investigated by simplifying the flow in both porous and non-porous media into a 2D problem.

More recently, \citet{howison_2000} and \citet{praud_2005} provided a comprehensive description of the mathematical formulation of immiscible two-phase flows in Hele-Shaw cells (also know as Saffman $\&$ Taylor problem). For a Hele-Shaw cell of a given size the development flow only depends on the capillary number. Thus if \textit{$N_{c}$} is too high \citet{saffman_1959}, \citet{homsy_1987} and \citet{tabeling_1987} determined that the flow develops into a single steady-state finger which moves through the cell with constant velocity $U$.

 Under the assumption that fluids remain immiscible along an interface, surface tension plays an important role in determining the shape and progress of the fingers~\citep{howison_2000}. During the displacement of a fluid by a less viscous one, the expected uniform front \citep{buckley_1942,sheldon_1959} is perturbed leading to an uneven front with elongations of the outside edge of the fluid interface (Fig.~\ref{fig:simple_case}), often referred as \textit{fingers}. In homogeneous domains, fingers start to develop when the surface tension acting on the interface between the fluids exert an opposite force towards the change of shape of the interface. The interface becomes unstable and collapses, taking a curved shape \citep{homsy_1987, jackson_2017}.

@@

Traditional computational geo-fluid dynamics (CGFD) models often rely on fixed mesh with sufficient resolution throughout the domain to capture specific flow dynamics (\eg fluid instabilities, flow re-circulation, heat and mass transfers etc), however computational overhead may be prohibitive for simulations involving complex geometries and heterogeneous properties.  Mesh adaptivity methods have been extensively used by the CGFD community to help capturing detailed flow dynamics, compositional non-equilibrium fluid displacement and solid-fluid interactions \citep{pluszny_2007,pietro_2014,su_2016,melnikova_2016}. In these methods, the mesh grid is continuously modified (\ie adjustments of the number and distribution of the degrees of freedom to reduce solution error) to focus resolution where is necessary as the simulation evolves in time and space. There are four main families of mesh adaptivity methods: adaptive mesh refinement (AMR), edge and face element manipulations whereas keeping the order of the element basis function fixed (h-adaptivity), mesh deformation (r-adaptivity) and changes of element basis function order (p-adaptivity). Detailed description and analysis of mesh adaptivity methods are beyond the scope of this manuscript but can be readily found in \citet{lo_book} \citep[see also][]{plewa_book,frey_book}.

@@

Here, two numerical simulations were conducted with fixed and adaptive mesh resolutions to qualitatively assess model capability to capture fingers dynamics. In order to trigger the formation of fingers, regions with sharp permeability gradient were introduced. The computational domain, Fig.~\ref{fig:testcase_heter_domain}a, consists of a rectangular geometry fully saturated with a fluid (except by a squared region containing 50$\%$ of a second fluid -- wetting fluid phase, Fig.~\ref{fig:testcase_heter_domain}b) and divided into 5 regions with prescribed permeability distribution. A no-flux boundary condition was imposed across upper and lower borders, whilst pure wetting fluid phase $\left(\text{\ie} S_{1}=1\right)$ is driven into the domain from the left-hand side face at velocity $u_{1}=1$. Viscosity ratio was set to 10.

Figures \ref{fig:5regions_fixedmesh} and \ref{fig:5regions_adaptmesh} show fluid displacement in simulations performed with fixed and adaptive meshes. Both simulations started with a mesh of 13068 triangular (\PN[1]{2}) elements, however as flow dynamics evolve the number of elements of the simulation conducted with adaptive mesh oscillates from a minimum of 4400 to a maximum of 16430 (Fig.~\ref{fig:5regions_plottimenodeselements}). In both cases, fingers' development (formation, growth and coalescence) and fluid cross flows (through regions of distinct permeabilities at the top of the domain, $\mathbf{K}=\left\{2,3\right\}$) can be readily noticed. The simulation performed with a fixed and relatively fine mesh was able to capture the progressive development of fingers (as shown in Fig.~\ref{fig:5regions_AdaptZoom}b) prompted by sharp permeability gradient at the interface of regions with $\mathbf{K}=\left\{1,5\right\}$. In regions with no permeability gradient, formation and development of fingers (Fig.~\ref{fig:5regions_AdaptZoom}c) were not captured by the simulation conducted with fixed mesh. In both cases, dynamic adaptive mesh based on perturbation of phase saturations with imposed interpolation error estimate proved to be able to capture the onset of instabilities and fingers' development with relatively little computational overhead, as indicated in Fig.~\ref{fig:5regions_plottimenodeselements}.

@@

As outlined in Section \ref{section:ViscousInstabilities}, viscous flow instabilities are characterised by spontaneous formation and growth of fluid elongations, called viscous fingers. Geometrical evolution in time and space of these elongations provides good indication of instabilities at the fluid front interface. Figure \ref{fig:3DChannel_PermMesh} shows the permeability distribution for the simulations conducted in this Section, where pure (wetting) fluid is added into the domain with uniform velocity of 2.5 cm.s$^{-1}$. No-flux boundary conditions are applied to all borders, except at the inlet (left-hand side) and outflow (right-hand side) faces. Initially, the domain was partially saturated with non-wetting fluid $\left(S_{nw}(t=0)=0.80\right)$, which was displaced by the wetting fluid as the simulation progress. The domain (with 235242 \PNDG[1]{1} tetrahedral element-pairs) was designed to naturally capture preferential flow pathways through channels of large absolute permeability (Fig.~\ref{fig:3DChannel_PermMesh}).% as shown in Fig.~\ref{fig:3DChannel_planes}-\ref{fig:3DChannel_Isosurf}.

@@

A numerical simulation of the same problem (\ie same geometry, initial mesh, initial and boundary conditions) was also performed using adaptive mesh, Fig.~\ref{fig:3DChannel_sat_adapt}, with resolution ranging from 235242 to 643064 tetrahedral \PNDG[1]{1} element-pairs (Fig.~\ref{fig:3DChannel_severalfields_adapt}a). Distinct preferable flow pathways can be shown (Fig.~\ref{fig:3DChannel_severalfields_adapt}c), in particular through the central channel (see streamlines in Fig.~\ref{fig:3DChannel_severalfields_adapt}b) and at the top of the domain.

>>>>

\begin{abstract} 
\red{(1) Motivation (1-2 sentences);}
We examine the flow instabilities, \ie viscous fingering on the displacement of one fluid by a second, using a novel computational method for modeling multi-phase flows in heterogeneous porous media.

\red{(2) Main aim of the manuscript highlighting the novelty of the work (you must not use the word novelty ... this should be implicit, 2-6 sentences);}
Starting from modelling the Hele-Shaw experimental apparatus, a mathematically analogous process to the two-dimensional flow in porous media problem, the finger development as a function of mobility ratio (MR), different domain set-up under uniform or different permeabilities regions and mesh types is examined. A series of cases are listed, showing the validation of the method against experimental setup followed by a more complex 2D domain setup under different viscosity ratios and different mesh resolutions. Finally a 3D domain setup is displaying the difference between the preferential flow path and the fingering regimes that will occur in the domain.

\red{(3) summary of the main technologies/methods (2-5 sentences)}
For these cases mentioned above, a novel control finite element method (CVFEM) is used to discretize the governing equations that will solve the multi-fluid porous media flow model. The governing multiphase porous media flow equations are solved using an unstructured mesh by introducing adaptivity in later stages. Dynamic mesh adaptivity will not only help to better capture the development of finger but also allow the system to focus the computational effort where is needed, minimizing both the CPU time and the numerical diffusion.

\red{(4) summary of the main findings}
It was found that viscosity ratio along with the initial and boundary conditions, as well as the use of adaptivity on top of an unstructured mesh will allow the user to directly apply and examine a geological model accurately and without the need of upscaling. Properties such as porosity and adaptivity can be easily introduced in the model while the idea of the preferential flow bath can be distinguished from the finger formation mechanics, in particular the tip-splitting finger behaviour.

\end{abstract}

@@

\medskip
%Due to the heterogeneity of geological formations, detailed modelling and simulation of multiphase flows through porous media is crucial for design, control and prediction of the aforementioned processes. 
Description of physics and mechanisms of multiphase porous media flows were reported by \citet{wooding_1976} with focus on capillary pressure and flow regimes. A comprehensive review of force balances at the interface between immiscible fluids and resulting mechanisms for flow instabilities can be found in \citet{homsy_1987}. Flow instabilities can be classified as macroscopic and microscopic, the former is due to imposed boundary conditions, whereas the later is associated to local phenomena at the fluids interface (\eg Kelvin-Helmholtz and Saffman-Taylor instabilities \citep{saffman_1959}). This work focuses on flow instabilities in two-phase systems due to viscous and stress forces (often referred as viscous instabilities or viscous fingering, Section \ref{section:ViscousInstabilities}).
%Description of physics and mechanisms of multiphase flows were reported by \citet{wooding_1976} with focus on capillary pressure and flow regimes. A comprehensive review of mechanisms of fluid instabilities (\ie fingering) and forces acting on the interface between two immiscible fluids can be found in \citet{homsy_1987}. \citet{adler_1988} described both microscopic and macroscopic scale Darcy-equations.
% Nevertheless we should refer to the area of flow instabilities  as these, first described by \citet{saffman_1958} \red{(IS THIS TRUE? SAFFMANN MAY HAVE BEEN THE FIRST TO THROUGHLY DISCUSS VISCOUS INSTABILITIES, BUT THERE WERE OTHER SCIENTISTS THAT DESCRIBED FLUID INSTABILITIES BEFORE, E.G., RAYLEIGH(1896), TAYLOR (1955) ETC)}. 
%The first distinction should be made between microscopic and macroscopic instabilities. Microscopic instabilities describe phenomena occurring locally, for example, Kelvin-Helmholtz and Saffman-Taylor instabilities \citep{saffman_1959}. On the other hand, macroscopic instabilities involve two-phase flow system \citep{bottoni_1992}. This paper focuses on microscopic instabilities in two-phase systems that occurs due to viscous and stress forces, often referred as viscous instabilities or \textit{viscous fingering} (see section \ref{section:ViscousInstabilities}).

@@

\medskip
The displacement efficiency depends upon the ratio of the viscous to capillary forces or capillary number ($N_{c}$, see eq.\ref{eq:capillary_number}). 
%Flow instabilities will occur due to: pressure gradient along the interface of fluids; stress forces acting in opposite directions; mobility differences that may exceed critical limits and; heterogeneities of the domain (layering). 
When viscous force of the injected fluid overcomes the capillary force, hydrodynamic instabilities may occur, resulting in the collapse of the interface between fluids and start forming fingers.

\medskip
Viscous flow instabilities can be found across several disciplines and scales, from chemical separation processes to geological reservoir fluids. \citet{muskat_1934} investigated fluid flow in Hele-Shaw cells, \ie parallel flat plates separated by an infinitesimal gap, and the impact on the capillary number ($N_{c}$),
\begin{equation}
N_{c} = \frac{\mu U}{\gamma}.
\label{eq:capillary_number}
\end{equation}

\noindent In Eqn.~\ref{eq:capillary_number}, $\gamma$ is the surface tension and $U$ is the characteristic velocity of the moving interface. This experimental apparatus enables instabilities to be qualitatively investigated by simplifying the flow in both porous and non-porous media into a 2D problem.

\medskip
More recently, \citet{howison_2000} and \citet{praud_2005} provided a comprehensive description of the mathematical formulation of immiscible two-phase flows in Hele-Shaw cells (also know as Saffman $\&$ Taylor problem). For a Hele-Shaw cell of a given size the development flow only depends on the capillary number. Thus if \textit{$N_{c}$} is too high \citet{saffman_1959}, \citet{homsy_1987} and \citet{tabeling_1987} determined that the flow develops into a single steady-state finger which moves through the cell with constant velocity $U$.

\medskip
 Under the assumption that fluids remain immiscible along an interface, surface tension plays an important role in determining the shape and progress of the fingers~\citep{howison_2000}. During the displacement of a fluid by a less viscous one, the expected uniform front \citep{buckley_1942,sheldon_1959} is perturbed leading to an uneven front with elongations of the outside edge of the fluid interface (Fig.~\ref{fig:simple_case}), often referred as \textit{fingers}. In homogeneous domains, fingers start to develop when the surface tension acting on the interface between the fluids exert an opposite force towards the change of shape of the interface. The interface becomes unstable and collapses, taking a curved shape \citep{homsy_1987, jackson_2017}.

@@

\medskip
Traditional computational geo-fluid dynamics (CGFD) models often rely on fixed mesh with sufficient resolution throughout the domain to capture specific flow dynamics (\eg fluid instabilities, flow re-circulation, heat and mass transfers etc), however computational overhead may be prohibitive for simulations involving complex geometries and heterogeneous properties.  Mesh adaptivity methods have been extensively used by the CGFD community to help capturing detailed flow dynamics, compositional non-equilibrium fluid displacement and solid-fluid interactions \citep{pluszny_2007,pietro_2014,su_2016,melnikova_2016}. In these methods, the mesh grid is continuously modified (\ie adjustments of the number and distribution of the degrees of freedom to reduce solution error) to focus resolution where is necessary as the simulation evolves in time and space. There are four main families of mesh adaptivity methods: adaptive mesh refinement (AMR), edge and face element manipulations whereas keeping the order of the element basis function fixed (h-adaptivity), mesh deformation (r-adaptivity) and changes of element basis function order (p-adaptivity). Detailed description and analysis of mesh adaptivity methods are beyond the scope of this manuscript but can be readily found in \citet{lo_book} \citep[see also][]{plewa_book,frey_book}.

@@

\medskip
Here, two numerical simulations were conducted with fixed and adaptive mesh resolutions to qualitatively assess model capability to capture fingers dynamics. In order to trigger the formation of fingers, regions with sharp permeability gradient were introduced. The computational domain, Fig.~\ref{fig:testcase_heter_domain}a, consists of a rectangular geometry fully saturated with a fluid (except by a squared region containing 50$\%$ of a second fluid -- wetting fluid phase, Fig.~\ref{fig:testcase_heter_domain}b) and divided into 5 regions with prescribed permeability distribution. A no-flux boundary condition was imposed across upper and lower borders, whilst pure wetting fluid phase $\left(\text{\ie} S_{1}=1\right)$ is driven into the domain from the left-hand side face at velocity $u_{1}=1$. Viscosity ratio was set to 10.

\medskip
Figures \ref{fig:5regions_fixedmesh} and \ref{fig:5regions_adaptmesh} show fluid displacement in simulations performed with fixed and adaptive meshes. Both simulations started with a mesh of 13068 triangular (\PN[1]{2}) elements, however as flow dynamics evolve the number of elements of the simulation conducted with adaptive mesh oscillates from a minimum of 4400 to a maximum of 16430 (Fig.~\ref{fig:5regions_plottimenodeselements}). In both cases, fingers' development (formation, growth and coalescence) and fluid cross flows (through regions of distinct permeabilities at the top of the domain, $\mathbf{K}=\left\{2,3\right\}$) can be readily noticed. The simulation performed with a fixed and relatively fine mesh was able to capture the progressive development of fingers (as shown in Fig.~\ref{fig:5regions_AdaptZoom}b) prompted by sharp permeability gradient at the interface of regions with $\mathbf{K}=\left\{1,5\right\}$. In regions with no permeability gradient, formation and development of fingers (Fig.~\ref{fig:5regions_AdaptZoom}c) were not captured by the simulation conducted with fixed mesh. In both cases, dynamic adaptive mesh based on perturbation of phase saturations with imposed interpolation error estimate proved to be able to capture the onset of instabilities and fingers' development with relatively little computational overhead, as indicated in Fig.~\ref{fig:5regions_plottimenodeselements}.

@@

\medskip
As outlined in Section \ref{section:ViscousInstabilities}, viscous flow instabilities are characterised by spontaneous formation and growth of fluid elongations, called viscous fingers. Geometrical evolution in time and space of these elongations provides good indication of instabilities at the fluid front interface. Figure \ref{fig:3DChannel_PermMesh} shows the permeability distribution for the simulations conducted in this Section, where pure (wetting) fluid is added into the domain with uniform velocity of 2.5 cm.s$^{-1}$. No-flux boundary conditions are applied to all borders, except at the inlet (left-hand side) and outflow (right-hand side) faces. Initially, the domain was partially saturated with non-wetting fluid $\left(S_{nw}(t=0)=0.80\right)$, which was displaced by the wetting fluid as the simulation progress. The domain (with 235242 \PNDG[1]{1} tetrahedral element-pairs) was designed to naturally capture preferential flow pathways through channels of large absolute permeability (Fig.~\ref{fig:3DChannel_PermMesh}).% as shown in Fig.~\ref{fig:3DChannel_planes}-\ref{fig:3DChannel_Isosurf}.

@@

\medskip
A numerical simulation of the same problem (\ie same geometry, initial mesh, initial and boundary conditions) was also performed using adaptive mesh, Fig.~\ref{fig:3DChannel_sat_adapt}, with resolution ranging from 235242 to 643064 tetrahedral \PNDG[1]{1} element-pairs (Fig.~\ref{fig:3DChannel_severalfields_adapt}a). Distinct preferable flow pathways can be shown (Fig.~\ref{fig:3DChannel_severalfields_adapt}c), in particular through the central channel (see streamlines in Fig.~\ref{fig:3DChannel_severalfields_adapt}b) and at the top of the domain.
